%Acknowledgments

\renewcommand{\baselinestretch}{2}
\small\normalsize
\hbox{\ }
 
\vspace{-.65in}

\begin{center}
\large{Acknowledgments} 
\end{center} 

\vspace{1ex}

First, I would like to express my gratitude to my advisor, Frauke Kreuter, for her guidance and mentorship. She was the first to suggest that I pursue a PhD in survey methodology, and our work together has helped me become a better scholar, speaker, and research professional. 
I wish to thank my committee -- Scott Keeter, Mike Brick, and Mike Elliott -- whose help, suggestions, and feedback were always on point. The diversity of their perspectives helped shape this work into something both rigorous and practical. 
Liz Stuart also deserves special thanks for sharing her expertise in causal inference and helping turn Chapter 2 into a scholarly publication.
I am grateful to my colleagues at Pew Research Center, especially Courtney Kennedy, whose support both in terms of data and time made this dissertation possible. 
I am similarly grateful to the faculty, staff, and fellow students at the Joint Program in Survey Methodology for everything I have learned, and for the community of survey researchers that has developed in JPSM’s orbit over the years. 
To my parents and siblings, I am thankful for their unwavering confidence in my eventual success.
Most of all, I am grateful to Kris, my partner in everything, who’s unflagging support and encouragement kept me going through the long and sometimes overwhelming task of completing this dissertation. This accomplishment would not have been possible without her.



